---
title: "PLANDOCDAY4"
---

## DAY4

------------------------------------------------------------------------

Short concept Time - Variability: 50 minutes

**Objectives:**

> Students know how to identify the measure of central tendency of a distribution, visually
>
> Students can calculate standard deviation from variance, and variance from standard deviation
>
> Students can look at two graphs and estimate which one has a higher variance

**Concepts:**

> Review: relative frequency histograms
>
> Review: probability distributions
>
> Review: shape
>
> Review: Central Tendencies (visual review)
>
> Variability: variance and stdev
>
> outliers

**Intuition Tools:**

> Dot counting
>
> Variability as big boot

**Lesson Material:**

Most of these concepts will be review, so I will quickly elaborate:

MEASURES OF CENTRAL TENDENCY: mean, median, mode

Then we will show a graph with a visible outlier, except we will note that outliers are somewhat subjective and not a mathematical truth, and that studies rarely just remove outliers as outliers can be expected

OUTLIER: a data point that differs significantly from the rest of the data. What counts as an outlier is subjective and outliers aren't usually removed from data outside of specific circumstances.

We will then go over some visual examples of higher variability in distributions, and that variacne and standard deviation are a means to measure this variability.

We will then state that the stdeviation is just the square root of the variance

$StandardDeviation=\sqrt{Variance}$

It will be noted that variance is calculated by taking the average difference between all of the values and the mean, so variance is another form of average, but not a measure of central tendency

NOTE: xbar=mean

$Variance(population)= \frac{\sum_{i=1}^{n}(x_{i}-\bar{x})}{n}$

$Variance(sample)= \frac{\sum_{i=1}^{n}(x_{i}-\bar{x})}{n-1}$

This formula will be of conceptual importance, but will likely not be actually used.

I will then show some examples of charts with low variance and charts with high variance (these are placeholders that were used in the handout, for lecture, I would prefer more chaotic sim examples)

```{r}
x<- seq(-3,3, length=10000)
y=dnorm(x, sd=0.5)

plot(x,y, main="distribution with lower variance")
```

```{r}
x<- seq(-3,3, length=10000)
y=dnorm(x, sd=1.5)

plot(x,y,
     ylim=c(0,.8),
     main="distribution with higher variance")
```

One way to think about this is to pretend that the distribution is being stomped on by a big boot, and the the higher the variance, the heavier the boot, the flatter and more spread it ends up

**ASSESSMENT:**

A single activity will be given to asses the concepts above. Students will be paired into groups with large writing devices. They will then be given a two large dot histogram plots on the screen and asked to identify the following

-   measures of central tendency (all 3)

-   identify any outliers

-   state which distribution has the higher variance

-   why, in words, you chose the distribution you chose.
